---
title: "Peptide Assay — Exploratory Analysis and Statistical Report"
subtitle: "Project 2 reference report (R · knitr)"
format: html
engine: knitr
execute:
  echo: true
  warning: false
---

## Setup

```{r}
#| label: setup
library(tidyverse)
library(broom)
set.seed(42)

dat <- read_csv("data/peptides_assay_clean.csv", show_col_types = FALSE)
dim(dat)
```

120 wells, 7 columns — the tidy table produced by Project 1.

## Exploratory analysis

```{r}
#| label: missingness
colSums(is.na(dat))
```

Only `concentration_ng_ul` has missing values (13 of 120). A missing
concentration means the *measurement* failed — bad well, pipetting error,
instrument hiccup — not that the sample didn't exist. That distinction
decides what we do with them: keep the rows, exclude per-analysis
(`na.rm`/`filter`), and ask in Q3 whether the failures cluster by batch.

```{r}
#| label: fig-raw
#| fig-width: 6
#| fig-height: 4
ggplot(dat, aes(concentration_ng_ul)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(x = "concentration (ng/uL)", y = "wells",
       title = "Concentrations pile up low with a long right tail")
```

```{r}
#| label: normality
shapiro.test(dat$concentration_ng_ul)$p.value
shapiro.test(log10(dat$concentration_ng_ul))$p.value
```

Raw-scale Shapiro rejects normality decisively (p ≈ 6×10⁻⁸) — the
histogram's skew is real, not sampling noise. On the log10 scale the test
finds no evidence against normality (p ≈ 0.73): **the data are
lognormal**, so all parametric testing happens on the log scale, and
medians (not means) summarize the raw values.

## Q1 — does treatment change concentration?

```{r}
#| label: q1
dat |>
  group_by(treatment) |>
  summarise(median_conc = median(concentration_ng_ul, na.rm = TRUE),
            n = sum(!is.na(concentration_ng_ul)), .groups = "drop")

t.test(concentration_ng_ul ~ treatment, data = dat)
t.test(log10(concentration_ng_ul) ~ treatment, data = dat) |>
  tidy()
```

Raw-scale Welch: |t| ≈ 0.06, p ≈ 0.96. Log-scale Welch: |t| ≈ 0.19,
p ≈ 0.85. Medians differ by ~8% in *either* direction of noise. **No
detectable treatment effect** — and per Q4 below, this design would have
caught a +50% effect most of the time, so this null is informative.

## Q2 — two-way ANOVA

```{r}
#| label: q2
dat |>
  filter(!is.na(concentration_ng_ul)) |>
  mutate(log_conc = log10(concentration_ng_ul)) |>
  aov(log_conc ~ treatment * peptide, data = _) |>
  tidy()
```

Every term is null: treatment p ≈ 0.81, peptide p ≈ 0.85, interaction
p ≈ 0.19. In words: no treatment effect, no peptide-to-peptide difference,
and — via the interaction — no evidence that treatment acts differently on
different peptides.

## Q3 — is missingness associated with batch?

```{r}
#| label: q3
tab <- table(dat$batch, is.na(dat$concentration_ng_ul),
             dnn = list("batch", "missing"))
tab
chisq.test(tab)
```

χ² ≈ 2.76, df = 2, p ≈ 0.25 — no association between batch and failure.
Honest caveat: the expected count in the "missing" cells is ≈ 4.3, below
the χ² rule-of-thumb of 5 (R warns about exactly this), so the exact test
is the safer word:

```{r}
#| label: q3-fisher
fisher.test(tab, workspace = 2e8)$p.value
```

Fisher's exact p ≈ 0.29 — same conclusion, sturdier footing.

## Q4 — would this design have detected a +50% effect?

```{r}
#| label: q4
n <- 50
pvals <- replicate(500, {
  t.test(rlnorm(n, log(25) + 0.4, 0.6),   # treated: +50%
         rlnorm(n, log(25),      0.6))$p.value
})
mean(pvals < 0.05)
```

Empirical power ≈ 0.85 (seeded, reproducible). If treatment truly boosted
concentrations by 50%, this design would have flagged it ~85% of the time
at these sample sizes. The Q1 null therefore means "no effect anywhere
near +50%", not "we couldn't see".

## Figures

```{r}
#| label: figs-box
#| fig-width: 5
#| fig-height: 4
dat |>
  filter(!is.na(concentration_ng_ul)) |>
  mutate(log_conc = log10(concentration_ng_ul)) |>
  ggplot(aes(treatment, log_conc, fill = treatment)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.4) +
  labs(x = NULL, y = "log10 concentration (ng/uL)",
       title = "Treatment groups are indistinguishable on the log scale")
```

```{r}
#| label: figs-int
#| fig-width: 6
#| fig-height: 4
dat |>
  filter(!is.na(concentration_ng_ul)) |>
  mutate(log_conc = log10(concentration_ng_ul)) |>
  group_by(treatment, peptide) |>
  summarise(m = mean(log_conc), .groups = "drop") |>
  ggplot(aes(peptide, m, color = treatment, group = treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = NULL, y = "mean log10 concentration",
       title = "Interaction plot: near-parallel lines agree with the ANOVA")
```

## Conclusion

The assay shows no treatment effect (Q1), no peptide differences, and no
treatment × peptide interaction (Q2), and measurement failures were random
with respect to batch (Q3). Because the design's power to detect a +50%
effect is ≈ 0.85 (Q4), these nulls are evidence of absence at that
effect size — not mere absence of evidence. The right next step is not
"run it again" but "define the smallest effect that matters"; if 20%
matters, Q4's simulation rerun at that effect size would prescribe a much
larger n before another plate is run.

## Reproducibility receipt

```{r}
#| label: receipt
format(Sys.time(), tz = "UTC", usetz = TRUE)
tools::md5sum("data/peptides_assay_clean.csv")
sessionInfo()
```

The Python twin (`projects/p2/solution.py`) reproduces every number in
this report — Q1 t/p to three decimals, the same ANOVA table, χ² = 2.76,
and the same power estimate — from the same seed. That cross-check is the
reproducibility thread: one analysis, two dialects, one receipt.
