---
title: "Peptide Assay — Exploratory Analysis and Statistical Report"
subtitle: "Project 2 · your name · the date you ran it"
format: html
engine: knitr
execute:
  echo: true
  warning: false
---

## Setup

Replace every `# TODO` below. Every random step uses the seed set here.

```{r}
#| label: setup
library(tidyverse)
library(broom)
set.seed(42)

dat <- read_csv("data/peptides_assay_clean.csv", show_col_types = FALSE)
dim(dat)
```

## Exploratory analysis

```{r}
#| label: missingness
# TODO: NAs per column (colSums(is.na(dat))). One sentence: what does a
# missing concentration MEAN physically?
```

```{r}
#| label: fig-raw
# TODO: histogram of concentration, raw scale — binwidth chosen on purpose,
# labeled axes, title that says the finding
```

```{r}
#| label: normality
# TODO: shapiro.test raw, then on log10(concentration_ng_ul).
# One sentence: which scale satisfies the test, and what does that make
# the data?
```

## Q1 — does treatment change concentration?

```{r}
#| label: q1
# TODO: Welch t-test on raw AND log10 scale (formula interface);
# medians by treatment; tidy() the log-scale test
```

## Q2 — two-way ANOVA

```{r}
#| label: q2
# TODO: aov(log10(conc) ~ treatment * peptide) on complete cases, tidy() it.
# One sentence per term: what does "null" mean here?
```

## Q3 — is missingness associated with batch?

```{r}
#| label: q3
# TODO: table(batch, is.na(conc)), chisq.test — and check the expected
# counts (they are < 5 in the 'missing' cells: what does that do to the
# approximation, and is fisher.test the safer answer?)
```

## Q4 — would this design have detected a +50% effect?

```{r}
#| label: q4
# TODO: 500 simulations, n = 50/group, rlnorm(log(25), 0.6) vs
# rlnorm(log(25) + 0.4, 0.6), Welch each time — fraction of p < 0.05
```

## Figures

```{r}
#| label: figs
# TODO: boxplot of log10 conc by treatment; interaction plot of mean
# log10 conc by peptide (one line per treatment). Day 3 pre-flight!
```

## Conclusion

TODO — five sentences maximum: what is null (Q1–Q3), what the power
estimate (Q4) does to that claim, and what you'd do next.

## Reproducibility receipt

```{r}
#| label: receipt
# TODO: Sys.time() (UTC), tools::md5sum of the data file, sessionInfo()
```
